---
title: "Regression I Final Project"
author: "Carina Doyle"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
rm (list = ls())
BMIdata <- read.table("PATH_W1.csv", header = TRUE, sep = ",")
```

BMI: body mass index (numerical)

Old: age range in years (1=15-17, 0=12-14)
Male: gender (1=male, 0=female)
Asthma: diagnosed (1=yes, 0=no)

MHint: mental health internalizing symptoms (1=yes, 0=no)
MHext: mental health externalizing symptoms (1=yes,0=no)

Race: race/ethnicity (1=hispanic, 2=white/reference, 3=black, 4=other)
Pedu: highest grade/school completed by parent (1=less than HS, 2=HS or equivalent, 3=some college or AA, 4=bachelor, 5=advanced)

Evertob: ever used any tobacco product (1=yes, 0=no)
Ftobacco: living with tobacco user (numerical)
NTobType: total number of tobacco type used (numerical)
SecondhandCig: number of hrs exposed (numerical)
Druguse: ever used (1=yes, 0=no)
Alcohol: ever used (1=yes, 0=no)
Marij: ever used (1=yes, 0=no)

Running descriptive statistics, checking for missing values
```{r}
library(Hmisc)
describe(BMIdata)
```
Recoding Race = White into reference group
```{r}
Path2 <- BMIdata
Path2$Race[Path2$Race==2] <- 0
```

Relabeling Old and Male
```{r}
Path2$Old <- factor(Path2$Old,labels = c("12-14 Years Old", "15-17 Years Old"))
Path2$Male <- factor(Path2$Male,labels = c("Male", "Female"))
```
Univariable Analysis 
DV: BMI
IV: Evertob + Old + Male + factor(Race)
```{r}
table(Path2$Evertob, exclude = NULL)

EvertobTAB <- table(Path2$Evertob)
prop.table(EvertobTAB)

summary(Path2$BMI)
sd(Path2$BMI)
IQR(Path2$BMI)

aggregate(BMI ~ Evertob, Path2, mean)
aggregate(BMI ~ Evertob, Path2, sd)
aggregate(BMI ~ Evertob, Path2, median)
aggregate(BMI ~ Evertob, Path2, IQR)
```

```{r}
table(Path2$Old)
OldTAB <- table(Path2$Old)
prop.table(OldTAB)

AgeByTobTAB <- table(Path2$Old, Path2$Evertob)
AgeByTobTAB
prop.table(AgeByTobTAB)
```

```{r}
table(Path2$Male)
MaleTAB <- table(Path2$Male)
prop.table(MaleTAB)

GenderByTobTAB <- table(Path2$Male, Path2$Evertob)
GenderByTobTAB
prop.table(GenderByTobTAB)
```

```{r}
table(Path2$Race)
RaceTAB <- table(Path2$Race)
prop.table(RaceTAB)

RaceByTobTAB <- table(Path2$Race, Path2$Evertob)
RaceByTobTAB
prop.table(RaceByTobTAB)
```

```{r}
table(Path2$MHint)
MHintTAB <- table(Path2$MHint)
prop.table(MHintTAB)

MHintByTobTAB <- table(Path2$MHint, Path2$Evertob)
MHintByTobTAB
prop.table(MHintByTobTAB)
```

```{r}
table(Path2$MHext)
MHextTAB <- table(Path2$MHext)
prop.table(MHextTAB)

MHextByTobTAB <- table(Path2$MHext, Path2$Evertob)
MHextByTobTAB
prop.table(MHextByTobTAB)
```

Bivariable Analysis
DV: BMI - continuous
IV: Evertob - dichotomous, Old - dichotomous, Male - dichotomous, Race - polytomous, MHint - dichotomous, MHext - dichotomous
```{r}
attach(Path2)
agelm <- lm(BMI~Old)
summary(agelm)
confint(agelm, 'Old', level=0.95)

evertoblm <- lm(BMI~Evertob)
summary(evertoblm)
confint(evertoblm, 'Evertob', level=0.95)

racelm <- lm(BMI~factor(Race))
summary(racelm)
confint(racelm, c('factor(Race)1','factor(Race)3', 'factor(Race)4') , level=0.95)

malelm <- lm(BMI~Male)
summary(malelm)
confint(malelm, 'Male', level=0.95)

intlm <- lm(BMI~MHint)
summary(intlm)
confint(intlm, 'MHint', level=0.95)

extlm <- lm(BMI~MHext)
summary(extlm)
confint(extlm, 'MHext', level=0.95)

detach(Path2)
```

---Model 1---
```{r}
model1 <- lm(BMI ~ Evertob + Old + Male + factor(Race), data=Path2)
summary(model1)
```

#Assumption 1- Linearity 
##Since the IVs are categorical, there is no need to establish linearity between IV and DV. 

#Assumption 2 - Independence of residuals 
```{r}
library(lmtest)
dwtest(model1)
```
#Assumption 3 - Normality 
```{r}
par (mfrow = c (1,2)) #standardized residuals vs fitted plot
qqnorm(rstandard(model1), main = "Q-Q plot")
abline(0,1,col = "blue")
hist(rstandard(model1), main = " ")

plot(model1,which=2)
```

#Assumption 4 - Equal Variance
```{r}
plot(model1, which = 3)

lmodResFit <- lm(sqrt(abs(residuals(model1))) ~ fitted(model1))
summary(lmodResFit)
```

#Assumption 5 - No Influential outliers
```{r}
library(faraway)
cd <- cooks.distance(model1)
n <- nrow(Path2)
p <- 4
cd[which(abs(cd)>4/n)]
halfnorm(cd, main= "Halfnorm plot for Cook's Distance")
```

#Assumption 6 - No Strong Multicollinearity 
```{r}
library(car)
car::vif(lm(BMI~Evertob+Old+Race, data=Path2)) #Assumption 6 not violated
```

---Model 2a - moderated effect of tobacco use on BMI by gender

Creating a model with interaction between tobacco use and gender
```{r}
model2a <- lm(BMI ~ Evertob + Old + Male + factor(Race) + Evertob*Male, data=Path2)
summary(model2a)
```

Checking significance of interaction term/Likelihood ratio of models
```{r}
anova(model1, model2a)
```

---Model 2b - moderated effect of tobacco use on BMI by age group

Creating a model with interaction between tobacco use and age group
```{r}
model2b <- lm(BMI ~ Evertob + Old + Male + factor(Race) + Evertob*Old, data=Path2)
summary(model2b)
```

Checking significance of interaction term/Likelihood ratio of models
```{r}
anova(model1, model2b)
```

---Model 3---

Creating a model including internal and external mental health symptoms
```{r}
model3a <- lm(BMI ~ Evertob + Old + Male + factor(Race) + Evertob + MHint + MHext, data=Path2)
summary(model3a)
```

Creating a model with gender interacting with tobacco use, internal and external mental health symptoms
```{r}
model3b <- lm(BMI ~ Evertob + Old + Male + factor(Race) + Evertob*Male + MHint*Male + MHext*Male, data=Path2)
summary(model3b)
```

Checking significance of interaction term/Likelihood ratio of models
```{r}
anova(model3a, model3b)
```
